---
title: "Appendix S3 - Group Sites by County and Flyway"
author: "Nicholas J Clark, David J Harris, Ceridwen Fraser"
output:
  pdf_document:
    highlight: zenburn
urlcolor: blue
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50),tidy=TRUE)
```

This appendix describes how to extract flyway and county grouping variables for each BBS survey site. These variables can be accessed in the `BBS.occurrences` library using `data('Site.descriptors')`.  
  
Load the cleaned BBS survey dataset that was prepared in `Appendix S1`
```{r eval = FALSE}
load("./BBS.Data.cleaned/BBS.2003-2009.filtered.Rdata")
```

Load the Landcover summary data for survey points that was prepared in `Appendix S2`
```{r eval = FALSE}
uniq.points.landcover <- read.csv("./LandCover/MODIS_Landcover_summary.csv")
```

Put landcover covariates and site descriptors into a single dataframe
```{r message=FALSE,warning=FALSE, eval = FALSE}
library(dplyr)
Site.descriptors = sites.df %>%
  dplyr::left_join(uniq.points.landcover, 
                   by = c('Latitude' = 'lat',
                          'Longitude' = 'long',
                          'year' = 'year')) %>%
  dplyr::select(-modis.band, -richness, -simpsons.d, -simpsons.evenness,
                -no.data.fill)
```

Here we write a function to group sites into counties. This is accomplished by loading the US Government Census `Cartographic Boundaries` shapefile. Note, this file can be manually downloaded here: https://www.census.gov/geo/maps-data/data/cbf/cbf_counties.html if the `tempfile` strategy below fails. We then use a simple function to ensure that the coordinate system of our latitudes and longitudes match that of the shapefile and then intersect each unique point with the counties
```{r eval = FALSE}
# install.packages('sf')
library(sf)
library(dplyr)
temp <- tempfile()
download.file('http://www2.census.gov/geo/tiger/GENZ2017/shp/cb_2017_us_county_5m.zip', temp)
county.shp <- sf::st_read(unzip(temp, 'cb_2017_us_county_5m.shp'))
unlink(temp)

geo_inside <- function(lon, lat, map, variable) {
#Escape the variable name
  variable <- dplyr::enquo(variable)
  
#Convert each unique point into an sf object
  pt <- tibble::data_frame(x = lon,
                           y = lat) %>%
    sf::st_as_sf(coords = c("x", "y"), 
                 crs = 4269) %>%
    sf::st_transform(crs = sf::st_crs(county.shp))
  
#Intersect the point with the map object
pt %>% 
  sf::st_join(map) %>% 
  dplyr::pull(!!variable)
}
```

\pagebreak

Next we write a similar function to group sites into regions based on recognised North American migratory flyways. This is accomplished by loading the US Fish and Wildlife Service's `WaterfowlFlyways` shapefile. Note, this file can be manually downloaded here: https://ecos.fws.gov/ServCat/Reference/Profile/42276 if the `tempfile` strategy below fails. We then use a simple function to ensure that the coordinate system of our latitudes and longitudes match that of the shapefile and then intersect each unique point with the flyway layer
```{r eval = FALSE}
# install.packages('sf')
library(sf)
library(dplyr)
temp <- tempfile()
download.file('https://ecos.fws.gov/ServCat/DownloadFile/113670', temp)
fly.shp <- sf::st_read(unzip(temp, 'WaterfowlFlyways.shp'))
unlink(temp)

geo_inside <- function(lon, lat, map, variable) {
#Escape the variable name
  variable <- dplyr::enquo(variable)
  
#Convert each unique point into an sf object
  pt <- tibble::data_frame(x = lon,
                           y = lat) %>%
    sf::st_as_sf(coords = c("x", "y"), 
                 crs = 4269) %>%
    sf::st_transform(crs = sf::st_crs(fly.shp))
  
#Intersect the point with the map object
pt %>% 
  sf::st_join(map) %>% 
  dplyr::pull(!!variable)
}
```

Now that all of our site-level covariates are stored together, we can intersect each unique point with the shapefile above to group the sites into counties.
```{r eval = FALSE}
point_counties = Site.descriptors %>%
  dplyr::select(Latitude, Longitude) %>%
  dplyr::distinct() %>%
  dplyr::mutate(County = geo_inside(lon = lon, 
                                         lat = lat,
                                  map = county.shp,
                                  variable = NAME))

Site.descriptors = Site.descriptors %>%
  left_join(point_counties, by = c('Latitude' = 'lat',
                                      'Longitude' = 'lon'))
```

Repeat for flyways and then join the grouping information back to the descriptors data
```{r eval = FALSE}
point_site.groups = Site.descriptors %>%
  dplyr::select(lat, lon) %>%
  dplyr::distinct() %>%
  dplyr::mutate(site.group = geo_inside(lon = lon, 
                                         lat = lat,
                                  map = fly.shp,
                                  variable = NAME))

Site.descriptors = Site.descriptors %>%
  left_join(point_site.groups, by = c('Latitude' = 'lat',
                                      'Longitude' = 'lon'))
colnames(Site.descriptors) <- stringr::str_to_title(colnames(Site.descriptors))
```

Save the grouped descriptor data for downstream analysis. Note, the flyway variables can be directly accessed in the `BBS.occurrences` library using `data('Site.descriptors')`.
```{r eval = FALSE}
dir.create('./Analysis_data')
save(Site.descriptors, file = './Analysis_data/Site.descriptors.Rdata')
```
